---
title: "Course Project 1"
author: "Josh"
date: "10/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

The first step in loading the data: 

```{r}
cls = c("integer", "character", "integer")
activity <- read.csv("activity.csv", head=TRUE, colClasses = cls, na.strings = "NA")
head(activity)
```

Next I will be changing up the data for later analysis:

```{r}
activity$date <- as.Date(activity$date)
activity_na <- subset(activity, !is.na(activity$steps))
```

## What is mean total number of steps taken per day?

Here I will be showing a histogram of the daily total number of steps taken:

```{r}
dailysum <- tapply(activity_na$steps, activity_na$date, sum, na.rm=TRUE, simplify=T)
dailysum <- dailysum[!is.na(dailysum)]

hist(x=dailysum,
     col="red",
     breaks=20,
     xlab="Daily total steps",
     ylab="Daily",
     main="The distribution of daily total")
```

Next, I will find the mean and median:

```{r}
mean(dailysum)
median(dailysum)
```

## What is the average daily activity pattern?

To find this, we create a time series plot (i.e. type = “l”) of the 5-minute interval and the average number of steps taken, averaged across all days:

```{r}
average_integer <- tapply(activity_na$steps, activity_na$interval, mean, 
                          na.rm=TRUE, simplify=T)
activity_integer_average <- data.frame(interval=as.integer(names(average_integer)), 
                                       avg=average_integer)

with(activity_integer_average,
     plot(interval,
                 avg,
                 type="l",
                 xlab="5-minute intervals",
                 ylab="Average Steps"))
```

Which of these contains the max number of steps?

``` {r}
max_steps <- max(activity_integer_average$avg)
activity_integer_average[activity_integer_average$avg == max_steps, ]
```

## Imputing missing values

Lets see how many there are in this dataset:

```{r}
sum(is.na(activity$steps))
```

We will create a new data frame with the missing data filled in. We will use the mean for that 5 minute interval.

```{r}
activity_impute <- activity
activitysteps_na <- is.na(activity_impute$steps)
integer_average <- tapply(activity_na$steps, activity_na$interval, mean, 
                          na.rm=TRUE, simplify=T)
activity_impute$steps[activitysteps_na] <- 
        integer_average[as.character(activity_impute$interval[activitysteps_na])]
```

Here is a histogram of the total steps per day:

```{r}
new_dailysum <- tapply(activity_impute$steps, activity_impute$date, sum, na.rm=TRUE, simplify=T)
hist(x=new_dailysum,
       col="red",
       breaks=20,
       xlab="Daily Steps",
       ylab="Daily",
       main="The distribution of daily total")
```

See below for the mean and median

```{r}
mean(new_dailysum)
median(new_dailysum)
```

## Are there differences in activity patterns between weekdays and weekends?

We'll first need to create a new factor variable "week" to assign each to weekday or weekend:

```{r}
is_weekday <- function(d) {
                wd <- weekdays(d)
                ifelse (wd == "Saturday" | wd == "Sunday", "weekend", "weekday")
        }

weekday_run <- sapply(activity_impute$date, is_weekday)
activity_impute$week <- as.factor(weekday_run)
head(activity_impute)
```

Next, we plot weekday and weekend to get a visual of the average number of steps taken:

```{r}
week_activity <- aggregate(steps ~ week+interval, data=activity_impute, FUN=mean)
library(lattice)
xyplot(steps ~ interval | factor(week),
         layout = c(1, 2),
         xlab="Interval",
         ylab="Number of steps",
         type="l",
         lty=1,
         data=week_activity)
```

There appears to be some differences in the two. The times they started to walk was much higher for weekday, while weekend has a more balanced start time.